Gastrointestinal III

### 2017-10-05 git3.rmd
sc

@@

>>>>

[Haemorrhoids](http://emedicine.medscape.com/article/775407-overview): *common* disease, varicose veins of perianal region. 
Causes = {pregnancy, constipation, obesity, portal hypertension, tumours}. 
Present: usually painless unless thrombosis, bleeding, "feel something when wiping". Rx: usually conservative (diet, stool softening), surgery: rubber band resection

-  Internal: above pectinate line, superior rectal
-  External: below pectinate line, inferior rectal, usually painful and usually thrombosed

Causes of GIT bleeding: split into upper, lower, anal (see: lecture 3)

Bowel cancer:

-  Left sided: constipation and is found earlier, blood on top of the poo (as poo has already formed)
-  Right sided: more liquidy so doesn't cause constipation, occult bleeding (blood mixed in)

Pain on defaecation: think haemorrhoids (thrombosed), anal fissure, anal polyps, abscesses, rectal ulcers, rectal cancer

[Laxatives](https://www.nps.org.au/australian-prescriber/articles/managing-constipation-in-adults):

-  Osmotic: indigestible carbohydrates => pass through => relative osmotic diarrhoea to soften stool
-  Bulking agents 
-  Stool softeners 
-  Stimulants
-  Lubricants

Diverticulosis: common, L in West, R in Asia, increases with age. 
Most are asymptomatic (70-80%), diverticular bleeding (outpouch with the vasa recta which is the weak part => bleeding not uncommon) (5-15%), diverticulitis (4%) = {simple (85%), complicated (15%)}

-  True: congenital, e.g. meckels
-  False: not all the layers, usually around vasa recta as that is the weakest, e.g. zenkers

Top *X* differentials, where $X=\in\mathbb{Z}$:

-  CRC: due to bleeding, and altered bowel habits
-  Diverticulitis/diverticulosis
-  Haemorrhoids

Explaining: emphasise importance and that it can cure things inside it

Further Ix:

-  Bloods: FBC (anaemia), ELFT, CRP/ESR, group+hold?, CEA (specific marker for Ca)
-  CT for staging

Risk factors for adenocarcinoma:

-  Smoking
-  Diet
-  Obesity
-  Genetic risk: FAP, lynch syndrome

Presentation: bleeding, abdo pain, tenesmus (feeling of incomplete emptying), anaemia (right), complete bowel obstruction,

Screening: important because catching it early has high survival rates

@@

DDx:

-  CVS: ruptured AAA => ischaemia of bowel, vasculitis
-  Renal: UTI + symptoms of cephalexin, pyelonephritis
-  GIT: ischaemic bowel, perforation, ulcer, CRC, diverticulitis, anal fissure, haemorrhoids, fistula, abscess, infectious colitis (*C. difficile* because in nursing home; everyone would get it)
-  Genitourinary: pelvic inflamm disease

Compared to Trevor (IBD, CRC), she is Old F in nursing home on antibiotics: infectious colitides, CRC (less likely presentation as her thing is so acute), diverticulitis, ischaemic bowel

Mechanisms that antibiotics causing diarrhoea:

-  Inflammatory: *C. difficile* secrete toxins
-  Secretory: damage mucosa
-  Osmotic: disturb carbohydrate balance
-  Motility: increase gut motility

She's a vasculopath

*Triple whammy for renal failure: ACE, diuretic, NSAID (aspirin)*

Ex:

-  CVS: important to check JVP
-  GIT: 5 Fs

Rigidity, guarding, rebound tenderness (pain after moving hand away after slowly palpating): suggests peritonitis/inflammation

Ix:

-  Bloods: FBC (anaemia), ELFT (kidney fn; she is on triple whammy for renal failure), ESR/CRP
-  ECG
-  Urine: urinalysis, dipstick
-  Stool: cultures (hard to grow *C. difficile*)
-  Imaging: CT (ischaemia less enhanced), angiograms
-  Colonoscopy

Ischaemic colitis (pain after eating; basically angina in GIT): global hypoperfusion => ischaemia. 
Can also be caused by strangulation, embolism, infn. 
Hers is an *acute on chronic* presentation.

Ischaemic colitis: chronic issue, usually resolves on its own, Rx: bowel rest, broad spectrum IV antibiotics
Acute mesenteric colitis: medical emergency
